Synthesis of Linear Functional Programs

### 2021-07-12 article/article.tex
tweaks

    % {TODO: Nao ha restrições sobre este? no código tinha que x não podia estar em
    % delta mas não faz muito sentido hah}
    \infer*[right=($\bang L$)]
    {\Gamma, y{:}A ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
    {\Gamma ; \Delta/\Delta ; \Omega, x{:}\bang A \Uparrow\ \vdash\
    \textrm{let}\ \bang y = x\ \textrm{in}\ M : C}

@@

The decision rules are naturally followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
%TODO: eu tenho uma assertion que garante que y não sai com Delta', mas agora
%parece-me inutil?
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
asynchronous. If we're left focused on an atomic proposition we either instance
the goal or fail. Otherwise the left focus is asynchronous and we can start
inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Ahead, we'll
present new rules that align and build on top of these to synth recursive
programs from more expressive (richer) types.

@@

The interesting challenge of polymorphism in synthesis is the usage of schemes
from the unrestricted context.  To begin with, $\Gamma$ now holds both types and
schemes. Consequently, after the rule \textsc{decideLeft!} is applied, we become
left focused on either a type or a scheme. Since left focus on a type is already
well formulated, only rules for left focus on a scheme are needed. Our algorithm
instances bound type vars of the focused scheme with existential type variables,
and the instanced type becomes the left focus. Inspired by the Hindley-Milner
system, we also generate inference constraints on the existential type variables
(postponing the decision of what type it should be to be used in the proof), and
collect them in a new constraints context (TODO: should be a substitution not a
context?) $\Theta$ that is propagated across derivation branches the same way
the linear context is (by havin an input and output context ($\Theta/\Theta'$)).
In contrast to HM, everytime a constraint is added it is solved against all
other constraints -- a branch of the proof search is desired to fail as soon as
possible; and more, two different existential vars do not unify against
one another ($?\alpha \mapsto ?\beta$ fails). Note that we instance the scheme with \emph{existential} type
variables ($\alpha$) rather than just type variables since the latter represent
universal types in the synth, and the first represent an existential instance of
a scheme, that might inclusively constraint normal type variables. Additionally,
it's required that all existential type variables were assigned a type, and
provided they were, we can forget the constraints generated from the usage of
the scheme. These concepts are formalized with the rules:
\begin{mathpar}
    \infer*[right=($\forall L$)]
    {
        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
        \\
        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
        \\
        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
    }
    {\Theta/\Theta; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
    \and
    \infer*[right=($?L$)]
    {\textsc{solve}(?\alpha
    \mapsto C, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
    x{:}?\alpha \Downarrow\ \vdash x : C}
    \and
    \infer*[right=($\Downarrow L?$)]
    {\textsc{solve}(?\alpha
    \mapsto A, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
    x{:}A \Downarrow\ \vdash x : ?\alpha}
\end{mathpar}
where
\begin{mathpar}
    \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'\ \textrm{means type}\
    \tau'\ \textrm{is an \emph{existential instantiation} of scheme}\ \forall
    \overline{\alpha}.\ \tau
    \and
    \textrm{ftv}_E(\tau')\ \textrm{is the set of free \emph{existential} type variables in type}\ \tau'
    \and
    ?\alpha \mapsto \tau_x\ \textrm{is a mapping from \emph{existential} type}\ ?\alpha\
    \textrm{to type}\ \tau_x
    \\
    \textsc{solve}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
    \textrm{can be solved with those in}\ \Theta
\end{mathpar}

>>>>

    \infer*[right=($\bang L$)]
    {\Gamma, y{:}A ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}\bang A \Uparrow\ \vdash\
    \textrm{let}\ \bang y = x\ \textrm{in}\ M : C}

@@

The decision rules are naturally followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
asynchronous. If we're left focused on an atomic proposition we either instance
the goal or fail. Otherwise the left focus is asynchronous and we can start
inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Ahead, we'll
present new rules that align and build on top of these to synth recursive
programs from more expressive (richer) types.

@@

The interesting challenge of polymorphism in synthesis is the usage of schemes
from the unrestricted context.  To begin with, $\Gamma$ now holds both types and
schemes. Consequently, after the rule \textsc{decideLeft!} is applied, we become
left focused on either a type or a scheme. Since left focus on a type is already
well formulated, only rules for left focus on a scheme are needed. Our algorithm
instances bound type vars of the focused scheme with existential type variables,
and the instanced type becomes the left focus. Inspired by the Hindley-Milner
system, we also generate inference constraints on the existential type variables
(postponing the decision of what type it should be to be used in the proof), and
collect them in a new constraints context $\Theta$ that is propagated across derivation branches the same way
the linear context is (by havin an input and output context ($\Theta/\Theta'$)).
In contrast to HM, everytime a constraint is added it is solved against all
other constraints -- a branch of the proof search is desired to fail as soon as
possible; and more, two different existential vars do not unify against
one another ($?\alpha \mapsto ?\beta$ fails). Note that we instance the scheme with \emph{existential} type
variables ($\alpha$) rather than just type variables since the latter represent
universal types in the synth, and the first represent an existential instance of
a scheme, that might inclusively constraint normal type variables. Additionally,
it's required that all existential type variables were assigned a type, and
provided they were, we can forget the constraints generated from the usage of
the scheme. These concepts are formalized with the rules:
\begin{mathpar}
    \infer*[right=($\forall L$)]
    {
        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
        \\
        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
        \\
        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
    }
    {\Theta/\Theta; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
    \and
    \infer*[right=($?L$)]
    {\textsc{solve}(?\alpha
    \mapsto C, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
    x{:}?\alpha \Downarrow\ \vdash x : C}
    \and
    \infer*[right=($\Downarrow L?$)]
    {\textsc{solve}(?\alpha
    \mapsto A, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
    x{:}A \Downarrow\ \vdash x : ?\alpha}
\end{mathpar}
where
\begin{mathpar}
    \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'\ \textrm{means type}\
    \tau'\ \textrm{is an \emph{existential instantiation} of scheme}\ \forall
    \overline{\alpha}.\ \tau
    \and
    \textrm{ftv}_E(\tau')\ \textrm{is the set of free \emph{existential} type variables in type}\ \tau'
    \and
    ?\alpha \mapsto \tau_x\ \textrm{is a mapping from \emph{existential} type}\ ?\alpha\
    \textrm{to type}\ \tau_x
    \\
    \textsc{solve}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
    \textrm{can be solved with those in}\ \Theta
\end{mathpar}
